% Created 2017-11-22 Wed 11:51
% Intended LaTeX compiler: pdflatex
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
%\usepackage{minted}
\input{preamble.tex}
\usepackage[backend=biber,firstinits=true,doi=false,isbn=false,url=false]{biblatex}
\addbibresource{refs.bib}
\date{December 2021}
\title{\vspace{-1cm}\myred{Refactoring}\vspace{-1cm}}
\begin{document}



\maketitle
\thispagestyle{empty}

\vspace{-1cm}
\begin{center}
\begin{tabular}{ll}
\gc candidate & Jakob Konrad Hansen\\
\gc supervisors & Volker Stolz and Lars Tveito\\
\gc type & 60ECTS\\
\gc study program & Informatics: Programming and System Architecture\\
\gc planned date of completion & May 2023\\
\end{tabular}
\end{center}

\vspace{-0.75cm}
\section*{Short description}
\label{sec:org4ecb37b}

The task is to implement and research refactoring for the ABS programming language\textsubscript{\cite{DBLP:conf/fmco/JohnsenHSSS10}}.

\vspace{-0.5cm}
\section*{Background and motivation}

Refactoring is the process of changing a software system in such a way that it preserves
the external behavior of the code, but also improves its internal
structure\textsubscript{\autocite{Fowler1999}}.


The thesis will be based around implementing refactoring support and researching
refactorings for the ABS programming language. We want to offer users of ABS more
refactorings and a satisfactory method to apply them. To do this we will build upon an
existing Xtext\textsubscript{\cite{xtext}} grammar and work with the LSP
protocol\textsubscript{\cite{lsp}}. A proposal to extend the LSP protocol might be made if
it aligns with the thesis. The temporary thesis statement and
research questions will be:

\vspace{-0.25cm}
\begin{itemize}
    \itemsep0em
    \item How should refactoring be implemented for ABS?
    \item Which refactorings can we safely implement for ABS?
    \item Is the resulting approach generalizable to other programming languages?
\end{itemize}
\vspace{-0.9cm}

\section*{Methodology}

The thesis will consist of both theoretical and practical work. The theoretical work will
consist of reading relevant research, formalizing refactorings, and argue their
correctness. The practical work will consist of implementing software which will provide
these refactorings. That will be a process of specifying requirements for the software,
implementing the software and then evaluating it.

\vspace{-0.5cm}
\section*{Progress plan}

\subsubsection*{Autumn 2021}
\begin{itemize}
    \itemsep0em
    \item Deliver the project plan and read relevant curriculum and papers
\end{itemize}
\subsubsection*{Spring 2022}
\vspace{-0.25cm}
\begin{itemize}
    \itemsep0em
    \item Read relevant papers
    \item Connect previously developed refactorings from Erlend Kristiansen's master
        thesis\textsubscript{\cite{DBLP:conf/nik/KristiansenS14}} with ABS, using LSP
    \item Narrow down the scope of the thesis based on the result of this implementation
    \item Formulate a thesis for and write essay
\end{itemize}
\vspace{-0.25cm}
\subsubsection*{Autumn 2022}
\vspace{-0.5cm}
\begin{itemize}
    \itemsep0em
    \item Analyze LSP and determine if the protocol is sufficient for refactoring
    \item Propose alternatives/extensions to the LSP protocol
    \item Implement refactorings for ABS 
\end{itemize}
\vspace{-0.25cm}
\subsubsection*{Spring 2023}
\vspace{-0.4cm}
\begin{itemize}
    \itemsep0em
    \item Evaluate refactorings and/or alternative LSP proposal
    \item Finish master thesis
\end{itemize}

\section*{Curriculum}

\subsection*{Relevant courses}
\begin{itemize}
    \item INF5110 - Kompilatorteknikk
    \item IN5440 - Selected topics in static analysis
    \item IN4240 - Testing av programvare
    \item IN5320 - Development in platform ecosystems
\end{itemize}

\subsection*{Relevant readings}

\begin{itemize}
    \item \emph{Refactoring and Active Object Languages}\textsubscript{\cite{DBLP:conf/isola/StolzPG20}}

    \item \emph{Refactoring}\textsubscript{\cite{Fowler1999}}
    \item \emph{Implementing Domain-Specific Languages with Xtext and Xtend}\textsubscript{\cite{lorenzoxtext}}
    \item \emph{The Specification Language Server Protocol: A Proposal for Standardised LSP Extensions}\textsubscript{\cite{DBLP:journals/corr/abs-2108-02961}}

\end{itemize}


\printbibliography
\end{document}

